---
title: "Assignment 3: Instrumental Variable Estimation"
subtitle: "Replicating the IV strategy in Stokes (2015)"
author: "EDS 241 / ESM 244 (DUE: 2/17/2026)"
format:
  html:
    theme: sketchy
    css: styles.css
date: "February 10, 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

------------------------------------------------------------------------

### Assignment instructions:

Working with classmates to troubleshoot code and concepts is encouraged. If you collaborate, list collaborators at the top of your submission.

All written responses must be written independently (in your own words).

Keep your work readable: Use clear headings and label plot elements thoughtfully (where applicable).

Submit both your rendered output and the Quarto (`.qmd`) file.

**Assignment submission (YOUR NAME)**: Isabella Segarra


------------------------------------------------------------------------

### Introduction

In this assignment, you will replicate the instrumental variable analysis from Stokes (2015), which examined how local wind turbine projects influenced electoral outcomes. Building on the matched dataset from Assignment 2, we will use Two-Stage Least Squares (2SLS) to estimate the causal effect of having a wind turbine proposed nearby on the change in Liberal Party vote share between 2007 and 2011. The instrument used in Stokes (2015) is a measure of local wind resource (average wind power, logged), which predicts where wind turbines are proposed. By using this instrument, we aim to isolate the portion of variation in turbine placement that is as-good-as-random, helping to meet the assumptions for causal identification.

**Study:** [Stokes, 2015 – Article](https://drive.google.com/file/d/1y2Okzjq2EA43AW5JzCvFS8ecLpeP6NKh/view?usp=drive_link)

**Data source:** [Dataverse – Stokes, 2015 replication data](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/SDUGCC)

------------------------------------------------------------------------

::: callout
Note: The estimates you obtain may not exactly match the published results in Stokes (2015) due to the alternative matching procedure used for processing the data in the previous assignment. Estimates should approximate the findings reported in *Table 2* of the article.
:::

------------------------------------------------------------------------

### Load packages

```{r}
library(tidyverse)
library(janitor)
library(here)
library(jtools)   # for export_summs (pretty regression tables)
library(AER)      # for ivreg (2SLS estimation)

```

------------------------------------------------------------------------

### Load the matched dataset (from Assignment 2)

> The matched_data has been preprocessed by matching on key covariates (e.g. pretreatment home values, education, income, population density) to improve balance between treated and control precincts. We will now use this data for the IV analysis. Make sure to re-code the `precinct_id` variable as a `factor`. 

```{r}

matched_data <- read_csv(here("data", "matched_data_subset.csv")) %>% 
    # Recode column as factor 
    mutate(precinct_id = as_factor(precinct_id))

```

------------------------------------------------------------------------

### Part 1: IV Identification Rationale

Intuition for Using an Instrument: 

**Question 1:** After matching on observables, why might we still need to utilize an instrumental variable approach to identify the causal effect of turbine proposals on vote share? In other words, what potential issues remain that an IV method can help address in this context? Use specific examples from the study to illustrate threats to a causal interpretation, then explain how an IV approach is designed to mitigate those threats.

*Response:* Although the distributions of matched covariates is similar, the  instrumental variable can influence the location of turbines (X), limiting the direct causal relationship between wind turbines and vote share. This is due to the fact that wind turbines were not randomly placed, but placed in regions with high wind power and wind power affects vote share only through turbine placement. Some potential issues mentioned in this study were that some communities might be more inclined to have wind turbines if they were open to climate policy or communities might oppose them if they had less minority populations, and by including wind power, the author was able to eliminate these differences in communities. 

------------------------------------------------------------------------

### Part 2: Two-Stage Least Squares (2SLS) Step-Wise Implementation

#### 2A. First-Stage Estimation: Regress the treatment ($D$) on the instrument ($Z$)

$$D_i = \alpha_0 + \alpha_1 Z_i $$

a. Estimate the first-stage regression of the treatment on the instrument (with controls). Regress `proposed_turbine_3km` on `log_wind_power`.
b. Include the control variables used in Stokes (2015) for both stages: Distance to lakes, geographic coordinates (latitude & longitude) with their squares and interaction, plus district fixed effects. 
c. After running the first stage, report the F-statistic for the instrument.

```{r}
first_stage <- lm(proposed_turbine_3km ~ log_wind_power + 
                      mindistlake_sq +
                      long_sq:lat_sq +
                      district_id, data = matched_data)

export_summs(first_stage, digits = 3,
             model.names = c("First stage: Prpoposed Turbine 3km"),
             coefs = c("(Intercept)", "log_wind_power")) 

```

###  Testing Instrument Relevance

Check instrument strength (F-statistic)
```{r}
# Extract the F-statistic for instrument strength
first_stage_summary <- summary(first_stage)

first_stage_summary$fstatistic[1]

```

**Question 2A:** Based on the instrument relevance test reported in the study, would you conclude the instrument is strong enough to be credible? Explain what a weak instrument would mean in this setting: Specifically, what would it suggest about compliance with Ontario’s Green Energy Act policy? 

*Response:* The instrument is ~81.1 which indicates a strong, relevant instrument for this study. This means that windpower has a strong relationship on proposed turbines which makes sense since turbines are placed in regions with good wind power. If this instrument were weak, then it would suggest that proposed turbines were placed in regions based on external factors such as political connections and not wind power. 


------------------------------------------------------------------------

#### 2B. Second Stage Estimation 

#### Regress the outcome ($Y$) on the fitted values from the 1st stage ($\hat{X}_i$)

$$Y_i = \beta_0 + \beta_1 \hat{D}_i + \epsilon_i$$

a. Now estimate the second stage of the 2SLS. 
b. First, use the first-stage model to generate the predicted values of proposed_turbine_3km for each precinct (these are $\hat{D}_i$). 
c. Add these predicted values as a new column in matched_data (e.g. proposed_turbine_3km_HAT). 
d. Then, regress the outcome change_liberal (the change in Liberal vote share from 2007 to 2011) on the predicted treatment (proposed_turbine_3km_HAT), including the same controls and fixed effects as in the first stage. 
e. Fill in the code for these steps below to obtain the second-stage regression results.

Save predicted values $\hat{X}_i$ from first stage 
```{r}
# Generate predicted values 
matched_data$proposed_turbine_3km_HAT <- predict(first_stage)
```

#### Estimate the second-stage regression 

$$LiberalVoteShare_i = \beta_0 + \beta_1 \widehat{ProposedTurbine}_i + ControlVariables... +  \epsilon_i$$

```{r}
second_stage <- lm(change_liberal ~ proposed_turbine_3km_HAT + mindistlake_sq +
                      long_sq:lat_sq +
                      district_id, data = matched_data)


export_summs(second_stage, digits = 3, 
             model.names = c("Second stage: Change in Liberal Vote Share"),
             coefs = c("(Intercept)", "proposed_turbine_3km_HAT") ) 
```

#### Interpreting the 2SLS Estimate

**Question 2B:** Imagine you are explaining your 2SLS findings to a policymaker in Ontario. What does the estimated coefficient on `proposed_turbine_3km_HAT` imply about the electoral impact of a local wind turbine proposal (within 3 km) on liberal vote share?

Proposed wind turbines within 3km of cities reduced liberal vote share by 19 percentage point. Precincts exposed to proposed turbines shifted away from the Liberal Party after they implemented the Green Energy Act. This effect is for precincts that got turbines because of strong wind power, not other external political factors. 

**Question 2C:** Explain what it means that IV identifies a LATE in the context of the wind-turbine voting study. What specific subset of observations does the second-stage 2SLS estimate apply to, and what does it imply about interpretation and generalizability? 

*Response:* IV identities the LATE or local average treatment effect for those where wind power influenced turbine placement. The second stage 2SLS applies to only precincts where strong wind power was the deciding factor for placing wind turbines. This interpretation only applies to that subset of precincts, not those that would get wind turbines for other factors such as politically favorable locations. 


------------------------------------------------------------------------

### Part 3: IV Assumptions and Validity

#### Evaluate Instrument Validity 

**Question 3:** List the four key assumptions required for the IV strategy (2SLS) to identify a causal effect, and briefly explain what each one means in the context of this study. (Hint: think about what conditions a valid instrument must satisfy (relevance, exclusion,...)

*Response:* Consider that Z is wind power potential and D is turbine placement and Y is vote share. This case study uses wind power as its instrumental variable. This variable satisfies the following four key assumptions: 
1) Exclusion restriction: Wind power (Z) solely influences vote share (Y) through turbine placement (D)
2) Relevance: As shown in the `first_stage` f-statistic, windpower (Z) strongly influences turbine placement, this is known as relevance. 
3) Ignorability (random first stage): Wind power is a random process that is determined through physical processes, not external political factors. 
4) Monotonicity: As shown with the `second_stage` model, precincts that had strong wind power strongly influenced proposed wind turbines. This satisfies the montonomicity assumption meaning that there were no precincts less likely to get turbines when windpower is higher. 

------------------------------------------------------------------------

### Part 4: Estimate 2SLS using `AER::ivreg()` 

📜 [SEE Documentation for specification details: AER package Viggnette Example](https://cran.r-project.org/web/packages/AER/vignettes/AER.pdf)

::: callout-tip

Syntax for specifying 2SLS using `ivreg()`:

`ivreg( Y ~ D + CONTROLS  | Z + CONTROLS , data )`

- The first-stage predictor variables go after the `~` symbol
- The second-stage predictor variables go after the `|` symbol

::: 


```{r}
  
fit_2sls <- ivreg(change_liberal ~ proposed_turbine_3km +
                      mindistlake_sq +
                      long_sq:lat_sq +
                      district_id |
                      log_wind_power +
                      mindistlake_sq +
                      long_sq:lat_sq +
                      district_id, 
                  data = matched_data
                      )

export_summs(fit_2sls, digits = 3,
             model.names = c("Change in Liberal Vote Share"),
             coefs = c("(Intercept)", "proposed_turbine_3km")
             ) 
```
#### Robustness checking strategies utilized in Stokes, 2015

**Question 4:** Choose two *robustness checks* from the paper that the authors use to increase confidence in their causal identification strategy. For each one, summarize the logic and findings from the robustness check in your own words:

*Response:* 
The author uses the instrumental variable as a robustness check to address selection bias. The IV is used in order to address endogeneity concerns or to address the fact that turbine placement is not random, but also determined by windpower which in itself, is separate from political factors. If the regression analysis shows that there is a relationship between windpower and turbine placement, then there is not external bias. Another robustness check the author utilized was within the proximity to wind turbines. The author looked at different distances to wind turbines to test whether the effect was proximity-based. This led them to use 3km as the preferred buffer. 

------------------------------------------------------------------------

### END

------------------------------------------------------------------------
